---
title: "Quantifying allelic bias"
author: "heinin"
date: "2023-12-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

### Packages and environment variables

```{r, results='hide', warning=F, message=F}

suppressPackageStartupMessages({library(cli)
                                library(Seurat)
                                library(SeuratObject)
                                library(SeuratDisk)
                                library(tidyverse)
                                library(tibble)
                                library(plyr)
                                library(dplyr)
                                library(ggplot2)
                                library(ggpubr)
                                library(ggrepel)
                                library(workflowr)
                                library(googlesheets4)
                                library(VariantAnnotation)})

setwd("/home/hnatri/ILD_ASE_Xenium/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/ILD_ASE_Xenium/code/colors_themes.R")
source("/home/hnatri/ILD_ASE_Xenium/code/plot_functions.R")
source("/home/hnatri/ILD_ASE_Xenium/code/utilities.R")

# Cell type annotations
gs4_deauth()
ct_annot  <- gs4_get("https://docs.google.com/spreadsheets/d/1SDfhxf6SjllxXEtNPf32ZKTEqHC9QJW3BpRYZFhpqFE/edit?usp=sharing")
sheet_names(ct_annot)
all_celltypes_annot <- read_sheet(ct_annot, sheet = "All celltypes, annotated, merged")

```

### Importing data

```{r}

seurat_object <- readRDS("/tgen_labs/banovich/IPF/Spatial_ASE/ILD_ASE_Xenium_annotated.rds")

```

### Importing eQTL and tag-SNP info

```{r}

gs4_deauth()
gtables  <- gs4_get("https://docs.google.com/spreadsheets/d/1f7HtQC07PWa1ij9UxTmgrEMX2OnHDVcrpn7Oa_VNIms/edit?usp=sharing")
sheet_names(gtables)
ld_snp_info <- read_sheet(gtables, sheet = "More candidates, SNP info with exon number")
ld_snp_info <- as.data.frame(ld_snp_info)

# Adding probe names
ld_snp_info$probe <- paste(ld_snp_info$gene, ld_snp_info$chr, ld_snp_info$linked_snp_pos, ld_snp_info$alt, sep = "-")
ld_snp_info$probe1 <- paste(ld_snp_info$gene, ld_snp_info$chr, ld_snp_info$linked_snp_pos, ld_snp_info$ref, sep = "-")
ld_snp_info$probe2 <- paste(ld_snp_info$gene, ld_snp_info$chr, ld_snp_info$linked_snp_pos, ld_snp_info$alt, sep = "-")

```

### Importing genotype data and creating SNPs matrices

```{r}

vcf <- readVcf("/tgen_labs/banovich/IPF/eQTL/qtl_mapping_vcf/filtered_MAF-HWE-INDPW.vcf", verbose = FALSE)
var_ranges <- rowRanges(vcf)
gt <- geno(vcf)
gt <- gt@listData$GT
snp_matrix <- genotypeToSnpMatrix(vcf)

rm(vcf)

# VCF metadata as a dataframe
gt_metadata <- cbind(as.data.frame(var_ranges@ranges),
                     data.frame("REF" = as.character(var_ranges@elementMetadata$REF),
                                "ALT" = as.character(unlist(var_ranges@elementMetadata$ALT))))
colnames(gt_metadata) <- c("var_start", "var_end", "var_width", "rsid", "REF", "ALT")

# Sanity check
ase_probes <- rownames(seurat_object)[grep("-", rownames(seurat_object))]

length(ase_probes)
length(ld_snp_info$probe1)
length(ld_snp_info$probe2)
length(intersect(c(ld_snp_info$probe1, ld_snp_info$probe2), ase_probes))
length(intersect(ld_snp_info$probe, ase_probes))
length(intersect(ld_snp_info$probe1, ase_probes))
length(intersect(ld_snp_info$probe2, ase_probes))

# Only keeping SNPs we need
gt_ase_snps <- gt[which(rownames(gt) %in% c(ld_snp_info$gene_body_snp, ld_snp_info$esnp)),]
rm(gt)
#write.csv(gt_ase_snps, "/scratch/hnatri/ILD/ILD_spatial_ASE/gt_ase_snps.csv")
#gt_ase_snps <- as.matrix(read.csv("/scratch/hnatri/ILD/ILD_spatial_ASE/gt_ase_snps.csv", row.names = "X"))

# Converting genotypes to dosage
gt_ase_snps <- gsub("0\\/0", 0, gt_ase_snps)
gt_ase_snps <- gsub("0\\/1", 1, gt_ase_snps)
gt_ase_snps <- gsub("1\\/1", 2, gt_ase_snps)
class(gt_ase_snps) <- "numeric"
gt_ase_snps <- as.data.frame(gt_ase_snps)

gt_eqtl_snps <- gt_ase_snps[which(rownames(gt_ase_snps) %in% ld_snp_info$esnp),]
gt_linked_snps <- gt_ase_snps[which(rownames(gt_ase_snps) %in% ld_snp_info$gene_body_snp),]

```

### Creating a dataframe with eSNP and tag-SNP genotype and allele info

```{r, results='hide', warning=F, message=F}
# Adding the probe name
gt_linked_snps$probe <- mapvalues(x = rownames(gt_linked_snps),
                                  from = ld_snp_info$gene_body_snp,
                                  to = ld_snp_info$probe)

# Pivot longer
gt_linked_snps <- gt_linked_snps %>% as.data.frame() %>%
  tidyr::pivot_longer(cols = setdiff(colnames(gt_linked_snps), c("probe")),
                      names_to = "Sample_Name",
                      values_to = "linked_SNP_GT") %>%
  mutate(gene = str_split(probe, "-", simplify = TRUE)[, 1],
         chr = str_split(probe, "-", simplify = TRUE)[, 2],
         pos = str_split(probe, "-", simplify = TRUE)[, 3],
         probe_var = str_split(probe, "-", simplify = TRUE)[, 4])

# Adding SNP info
gt_linked_snps$gene_body_snp <- mapvalues(x = gt_linked_snps$probe,
                                          from = ld_snp_info$probe,
                                          to = ld_snp_info$gene_body_snp)
gt_linked_snps$esnp <- mapvalues(x = gt_linked_snps$probe,
                                 from = ld_snp_info$probe,
                                 to = ld_snp_info$esnp)

# Adding eQTL genotypes
gt_eqtl_snps <- gt_eqtl_snps %>% as.data.frame() %>%
  rownames_to_column(var = "esnp") %>%
  tidyr::pivot_longer(cols = setdiff(colnames(gt_eqtl_snps), c("esnp")),
                      names_to = "Sample_Name",
                      values_to = "eSNP_GT")

# Merging eSNP and tag-SNP genotype and SNP info
gt_data <- merge(gt_linked_snps, gt_eqtl_snps, by = c("Sample_Name", "esnp"))

# Adding REF and ALT alleles
gt_data$tag_REF <- plyr::mapvalues(x = gt_data$gene_body_snp,
                                   from = gt_metadata$rsid,
                                   to = gt_metadata$REF)
gt_data$tag_ALT <- plyr::mapvalues(x = gt_data$gene_body_snp,
                                   from = gt_metadata$rsid,
                                   to = gt_metadata$ALT)
gt_data$eSNP_REF <- plyr::mapvalues(x = gt_data$esnp,
                                    from = gt_metadata$rsid,
                                    to = gt_metadata$REF)
gt_data$eSNP_ALT <- plyr::mapvalues(x = gt_data$esnp,
                                    from = gt_metadata$rsid,
                                    to = gt_metadata$ALT)

# A vector of all gene_tag_SNP -pairs
gene_tagSNPs <- unique(paste0(gt_linked_snps$gene, "_", gt_linked_snps$gene_body_snp))

```

### Obtaining pseudo-bulk expression values

```{r}

pseudobulk <- lapply(unique(seurat_object$annotation_3), function(ct){
  seurat_subset <- subset(seurat_object, subset = annotation_3 == ct)
  exp <- AverageExpression(seurat_object,
                           group.by = "Sample_Name",
                           assay = "RNA")
  
  as.data.frame(exp$RNA)
})

names(pseudobulk) <- unique(seurat_object$annotation_3)

```

### Looping through each cell type and gene-tag-SNP -pair and finding the 
allelic ratio

ASE is shown as the ALT/REF ratio of read counts.

```{r}

ld_snp_info$gene_tagSNP <- paste0(ld_snp_info$gene, "_", ld_snp_info$gene_body_snp)

ase_res <- lapply(names(pseudobulk), function(ct){
  # Expression data
  exp_data <- pseudobulk[[ct]]
  
  # Creating a dataframe for the results
  corr_res <- data.frame(matrix(NA, ncol = ncol(exp_data),
                                nrow = length(gene_tagSNPs)))
  rownames(corr_res) <- gene_tagSNPs
  colnames(corr_res) <- colnames(exp_data)
  
  # Looping through every locus, calculating the allelic ratio
  for(i in rownames(corr_res)){
    # Finding the probe names for REF and ALT
    locus_info <- ld_snp_info %>%
      filter(gene_tagSNP == i) %>%
      dplyr::select(-c("esnp", "gene_snp", "R2", "esnp_loc", "type")) %>%
      distinct()
    # Calculating the ratio
    ref_exp <- exp_data[locus_info$probe1,]
    alt_exp <- exp_data[locus_info$probe2,]
    corr_res[i,] <- alt_exp/ref_exp
  }
  
  corr_res$celltype <- ct
  
  corr_res
})

names(ase_res) <- names(pseudobulk)

```

### Applyin a Wilcoxon rank-sum test to calculate the statistical significance 
of the differential ASE between genotypes

```{r}

ase_sigtest <- lapply(names(ase_res), function(ct){
  # Allelic bias values
  ct_ase <- ase_res[[ct]]
  
  # Creating a dataframe for the results
  wilcox_res <- data.frame(matrix(NA, ncol = 4,
                                  nrow = length(gene_tagSNPs)))
  rownames(wilcox_res) <- gene_tagSNPs
  colnames(wilcox_res) <- c("p_2vs0", "p_1_2vs0", "p_adj_2vs0", "p_adj_1_2vs0")
  
  # Looping through every locus, calculating the allelic ratio
  for(i in rownames(wilcox_res)){
    tagSNP <- strsplit(i, split = "_")[[1]][2]
    
    # ASE values
    ct_ase_vals <- ct_ase %>% filter(row.names(ct_ase) == i) %>%
      dplyr::select(-c("celltype")) %>%
      t() %>% as.data.frame() %>%
      rownames_to_column(var = "Sample_Name")
    # Adding genotype information
    gt_vals <- gt_data %>% filter(gene_body_snp == tagSNP)
    # gt_vals[which(gt_vals$linked_SNP_GT==0), "Sample_Name"]
    ase_00 <- ct_ase_vals %>% filter(Sample_Name %in% gt_vals[which(gt_vals$linked_SNP_GT==0), "Sample_Name"])
    ase_01 <- ct_ase_vals %>% filter(Sample_Name %in% gt_vals[which(gt_vals$linked_SNP_GT==1), "Sample_Name"])
    ase_11 <- ct_ase_vals %>% filter(Sample_Name %in% gt_vals[which(gt_vals$linked_SNP_GT==2), "Sample_Name"])
    
    if(nrow(ase_11)==0 | nrow(ase_00)==0){
      next
    }
    if(length(unique(c(ase_00[,i], ase_01[,i], ase_11[,i])))==1){
      next
    }
    
    wres1 <- wilcox.test(c(ase_11[,i], ase_01[,i]), ase_00[,i], alternative = "two.sided")
    wres2 <- wilcox.test(ase_11[,i], ase_00[,i], alternative = "two.sided")
    
    wilcox_res[i, c("p_1_2vs0")] <- wres1[["p.value"]]
    wilcox_res[i, c("p_2vs0")] <- wres2[["p.value"]]
  }
  
  wilcox_res$celltype <- ct
  # Adjusting for multiple comparisons
  wilcox_res["p_adj_1_2vs0"] <- p.adjust(wilcox_res[,"p_1_2vs0"],
                                         n = length(wilcox_res[,"p_1_2vs0"]))
  wilcox_res["p_adj_2vs0"] <- p.adjust(wilcox_res[,"p_2vs0"],
                                   n = length(wilcox_res[,"p_2vs0"]))
  
  wilcox_res
})

names(ase_sigtest) <- names(ase_res)

# Saving the results for later
saveRDS(ase_sigtest, "/scratch/hnatri/ILD/ILD_spatial_ASE/ase_sigtest.rds")

```

### Testing between all groups using Kruskal–Wallis rank-sum test

```{r}

```




