---
title: "Lineage level annotations"
author: "heinin"
date: "2023-12-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Cell type annotations for spatial data.

### Packages and environment variables

```{r}

suppressPackageStartupMessages({library(cli)
                                library(Seurat)
                                library(SeuratObject)
                                library(SeuratDisk)
                                library(tidyverse)
                                library(tibble)
                                library(ggplot2)
                                library(ggpubr)
                                library(ggrepel)
                                library(workflowr)
                                library(googlesheets4)})

setwd("/home/hnatri/ILD_ASE_Xenium/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/ILD_ASE_Xenium/code/colors_themes.R")
source("/home/hnatri/ILD_ASE_Xenium/code/plot_functions.R")
source("/home/hnatri/ILD_ASE_Xenium/code/utilities.R")

```

### Import data

```{r}

seurat_object <- readRDS("/tgen_labs/banovich/IPF/Spatial_ASE/ILD_ASE_Xenium_processed_npcs20.rds")
DefaultAssay(seurat_object)

```

### Feature expression

```{r}

# Dimplot of clusters
DimPlot(seurat_object,
        group.by = "leiden_res0.5",
        cols = cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  NoLegend() +
  theme_minimal()

```

```{r, fig.height=12}

# Cell type markers
DotPlot(seurat_object,
        features = unique(c(epithelial_features,
                            immune_features,
                            endothelial_features,
                            mesenchymal_features)),
        group.by = "leiden_res0.5",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()

```

### Lineage annotations

```{r}
epi_clusters <- c(9, 7, 8, 9)
imm_clusters <- c(3, 5, 6, 10)
endo_mesen_clusters <- c(1, 2, 4)
```

## Epithelial
### Subsetting clusters with any *EPCAM* expression and reclustering

```{r}

epithelial <- subset(seurat_object, subset = leiden_res0.5 %in% epi_clusters)

# Saving the object for clustering with scanpy
#saveRDS(epithelial, "/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial.rds")
#epithelial <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_clustered.rds")

# Reclustering using Seurat. The function is sourced from /code/utilities.R
epithelial_reclustered <- recluster(epithelial)
#saveRDS(epithelial_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_reclustered.rds")

```

### Marker expression

```{r}
#epithelial_reclustered <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_reclustered.rds")

# Dimplot of clusters
DimPlot(epithelial_reclustered,
        group.by = "snn_res.0.3",
        cols = cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  NoLegend() +
  theme_minimal()
```

```{r, fig.height=12}
# Cell type markers
DotPlot(epithelial_reclustered,
        features = unique(c(epithelial_features,
                            immune_features,
                            endothelial_features,
                            mesenchymal_features)),
        group.by = "snn_res.0.3",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()
```

```{r, fig.height=8}
FeaturePlot(epithelial_reclustered,
            features = epithelial_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(epithelial_reclustered,
            features = immune_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(epithelial_reclustered,
            features = endothelial_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(epithelial_reclustered,
            features = mesenchymal_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

```

Some clusters are not epithelial:\
* Cluster 3: endothelial\
* Clusters 0, 4, 5, 6, 8, 9, 10: immune\

### Adding lineage to the metadata

```{r}
epithelial_reclustered$lineage <- ifelse(epithelial_reclustered$snn_res.0.3 %in% c(1, 2), "Epithelial",
                                  ifelse(epithelial_reclustered$snn_res.0.3 %in% c(3), "Endothelial",
                                  ifelse(epithelial_reclustered$snn_res.0.3 %in% c(0, 4, 5, 6, 7, 8, 9, 10), "Immune", NA)))

#saveRDS(epithelial_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_reclustered.rds")
```

## Immune
### Subsetting clusters with immune marker expression and reclustering

```{r}

immune <- subset(seurat_object, subset = leiden_res0.5 %in% imm_clusters)

# Saving the object for clustering with scanpy
#saveRDS(epithelial, "/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial.rds")
#epithelial <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_clustered.rds")

# Reclustering using Seurat. The function is sourced from /code/utilities.R
immune_reclustered <- recluster(immune)
#saveRDS(immune_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/immune_reclustered.rds")

```

### Marker expression

```{r}
#immune_reclustered <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/immune_reclustered.rds")

# Dimplot of clusters
DimPlot(immune_reclustered,
        group.by = "snn_res.0.3",
        cols = cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  NoLegend() +
  theme_minimal()
```

```{r, fig.height=12}
# Cell type markers
DotPlot(immune_reclustered,
        features = unique(c(epithelial_features,
                            immune_features,
                            endothelial_features,
                            mesenchymal_features)),
        group.by = "snn_res.0.3",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()
```

```{r, fig.height=8}
FeaturePlot(immune_reclustered,
            features = epithelial_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(immune_reclustered,
            features = immune_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(immune_reclustered,
            features = endothelial_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(immune_reclustered,
            features = mesenchymal_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

```

Some clusters are not immune:
* Cluster 5: epithelial
* Clusters 3, 4: endothelial
* Clusters

### Adding lineage to the metadata

```{r}
immune_reclustered$lineage <- ifelse(immune_reclustered$snn_res.0.3 %in% c(5), "Epithelial",
                              ifelse(immune_reclustered$snn_res.0.3 %in% c(3, 4), "Endothelial",
                              ifelse(immune_reclustered$snn_res.0.3 %in% c(0, 1, 2, 6, 7, 8), "Immune", NA)))

#saveRDS(immune_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/immune_reclustered.rds")
```

## Endothelial and mesenchymal
### Subsetting and reclustering

```{r}

endo_mesen <- subset(seurat_object, subset = leiden_res0.5 %in% endo_mesen_clusters)

# Saving the object for clustering with scanpy
#saveRDS(epithelial, "/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial.rds")
#epithelial <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_clustered.rds")

# Reclustering using Seurat. The function is sourced from /code/utilities.R
endo_mesen_reclustered <- recluster(endo_mesen)
#saveRDS(endo_mesen_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_reclustered.rds")

```

### Marker expression

```{r}
#endo_mesen_reclustered <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_reclustered.rds")

# Dimplot of clusters
DimPlot(endo_mesen_reclustered,
        group.by = "snn_res.0.3",
        cols = cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  NoLegend() +
  theme_minimal()
```

```{r, fig.height=12}
# Cell type markers
DotPlot(endo_mesen_reclustered,
        features = unique(c(epithelial_features,
                            immune_features,
                            endothelial_features,
                            mesenchymal_features)),
        group.by = "snn_res.0.3",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()
```

```{r, fig.height=8}
FeaturePlot(endo_mesen_reclustered,
            features = epithelial_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(endo_mesen_reclustered,
            features = immune_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(endo_mesen_reclustered,
            features = endothelial_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()

FeaturePlot(endo_mesen_reclustered,
            features = mesenchymal_features,
            ncol = 5,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  NoLegend() &
  theme_minimal()


```

Some clusters are not endothelial/mesenchymal:\
* Cluster 0: epithelial\
* Cluster 2, 4, 6, 7: immune\

### Adding lineage to the metadata

```{r}
endo_mesen_reclustered$lineage <- ifelse(endo_mesen_reclustered$snn_res.0.3 %in% c(0), "Epithelial",
                                  ifelse(endo_mesen_reclustered$snn_res.0.3 %in% c(2, 4, 6, 7), "Immune", "Endo_Mesen"))

#saveRDS(endo_mesen_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_reclustered.rds")
```

## Subsetting and merging at the lineage level

```{r}
unique(epithelial_reclustered$lineage)
unique(immune_reclustered$lineage)
unique(endo_mesen_reclustered$lineage)

epithelial_subsets <- list("epi1" = subset(epithelial_reclustered, subset = lineage == "Epithelial"),
                           "epi2" = subset(immune_reclustered, subset = lineage == "Epithelial"),
                           "epi3" = subset(endo_mesen_reclustered, subset = lineage == "Epithelial"))

immune_subsets <- list("imm1" = subset(epithelial_reclustered, subset = lineage == "Immune"),
                       "imm2" = subset(immune_reclustered, subset = lineage == "Immune"),
                       "imm3" = subset(endo_mesen_reclustered, subset = lineage == "Immune"))

endo_mesen_subsets <- list("em1" = subset(epithelial_reclustered, subset = lineage %in% c("Endo_Mesen", "Endothelial")),
                           "em2" = subset(immune_reclustered, subset = lineage %in% c("Endo_Mesen", "Endothelial")),
                           "em3" = subset(endo_mesen_reclustered, subset = lineage %in% c("Endo_Mesen", "Endothelial")))

epithelial_merged <- merge(x = epithelial_subsets[[1]],
                           y = epithelial_subsets[2:length(epithelial_subsets)])
immune_merged <- merge(x = immune_subsets[[1]],
                       y = immune_subsets[2:length(immune_subsets)])
endo_mesen_merged <- merge(x = endo_mesen_subsets[[1]],
                           y = endo_mesen_subsets[2:length(endo_mesen_subsets)])

# Saving
#saveRDS(epithelial_merged, "/scratch/hnatri/ILD/ILD_spatial_ASE/epithelial_merged.rds")
#saveRDS(immune_merged, "/scratch/hnatri/ILD/ILD_spatial_ASE/immune_merged.rds")
#saveRDS(endo_mesen_merged, "/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_merged.rds")

```


