---
title: "Annotating endothelial and mesenchymal cell types"
author: "heinin"
date: "2023-12-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## First pass annotations on the epithelial lineage

### Packages and environment variables

```{r}

suppressPackageStartupMessages({library(cli)
                                library(Seurat)
                                library(SeuratObject)
                                library(SeuratDisk)
                                library(tidyverse)
                                library(tibble)
                                library(plyr)
                                library(ggplot2)
                                library(ggpubr)
                                library(ggrepel)
                                library(workflowr)
                                library(googlesheets4)
                                library(presto)})

setwd("/home/hnatri/ILD_ASE_Xenium/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/ILD_ASE_Xenium/code/colors_themes.R")
source("/home/hnatri/ILD_ASE_Xenium/code/plot_functions.R")
source("/home/hnatri/ILD_ASE_Xenium/code/utilities.R")

# Cell type annotations
gs4_deauth()
ct_annot  <- gs4_get("https://docs.google.com/spreadsheets/d/1SDfhxf6SjllxXEtNPf32ZKTEqHC9QJW3BpRYZFhpqFE/edit?usp=sharing")
sheet_names(ct_annot)

```

### Import data

```{r}

endo_mesen_merged <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_merged.rds")
DefaultAssay(endo_mesen_merged)

unique(endo_mesen_merged@meta.data$Sample_Name)

# Adding one more cluster from the annotated immune object
immune_reclustered <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/immune_reclustered_annotated_orig.rds")

endo_mesen_merged <- merge(endo_mesen_merged,
                           subset(immune_reclustered,
                                  subset = annotation_1 %in% c("Endothelial", "Endo_Mesen", "Fibroblast")))

```

### Redonstructing the dimensionality reduction and neighborhood graph

```{r}

# Already ran this previously, using the same object
#endo_mesen_reclustered <- recluster(endo_mesen_merged)
endo_mesen_reclustered <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_reclustered_annotated.rds")

DimPlot(endo_mesen_reclustered,
        group.by = "snn_res.0.8",
        reduction = "umap",
        raster = T,
        cols = cluster_col,
        label = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()

DimPlot(endo_mesen_reclustered,
        group.by = "snn_res.0.8",
        split.by = "snn_res.0.8",
        reduction = "umap",
        ncol = 4,
        raster = T,
        cols = cluster_col,
        label = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()

# Numbers of cells per cluster
table(endo_mesen_reclustered$snn_res.0.8) %>% as.data.frame() %>%
  ggplot(aes(x = Var1, y = Freq)) +
  geom_bar(stat="identity", fill = "gray89") +
  xlab("snn_res.0.8") +
  ylab("# cells") +
  theme_minimal()

```

### Marker expression

```{r}
# Checking some lineage markers
FeaturePlot(endo_mesen_reclustered,
            features = c("EPCAM", "AGER", "CD3E", "CD19", "CD34", "SFRP2"),
            ncol = 3,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()
```

```{r, fig.height=12}
# All endothelial and mesenchymal markers
FeaturePlot(endo_mesen_reclustered,
            features = unique(c(endothelial_features, mesenchymal_features)),
            ncol = 4,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()
```

```{r, fig.height=10}
DotPlot(endo_mesen_reclustered,
        features = unique(c(endothelial_features, mesenchymal_features)),
        group.by = "snn_res.0.8",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()
```

### Top markers for each cluster

```{r}
# Only looking at base panel features
features <- rownames(endo_mesen_reclustered)
features <- features[-grep("-", features)]
base_counts <- LayerData(endo_mesen_reclustered,
                         assay = "RNA",
                         layer = "counts")
base_counts <- base_counts[features, ]

# Creating a new assay
# Ran this already, skipping
#endo_mesen_reclustered[["base_RNA"]] <- CreateAssay5Object(counts = base_counts)
#DefaultAssay(endo_mesen_reclustered) <- "base_RNA"
#endo_mesen_reclustered <- NormalizeData(endo_mesen_reclustered,
#                                        assay = "base_RNA",
#                                        normalization.method = "LogNormalize",
#                                        verbose = F)

# Comparing Seurat and presto
Idents(endo_mesen_reclustered) <- endo_mesen_reclustered$snn_res.0.8
cluster_markers <- FindAllMarkers(endo_mesen_reclustered,
                                  assay = "base_RNA",
                                  logfc.threshold = 0.25,
                                  test.use = "wilcox",
                                  slot = "data",
                                  min.pct = 0.1,
                                  verbose = F)
cluster_markers_presto <- presto::wilcoxauc(endo_mesen_reclustered,
                                            group_by = "snn_res.0.8",
                                            assay = "data",
                                            seurat_assay = "base_RNA")

# Overlap of top markers for cluster 0
cluster_markers_sig <- cluster_markers %>%
  filter(p_val_adj<0.01, abs(avg_log2FC)>0.5) %>%
  filter(cluster==0) %>%
  select(gene) %>% unlist() %>% as.character()
  
cluster_markers_presto_sig <- cluster_markers_presto %>%
  filter(padj<0.01, abs(logFC)>0.5) %>%
  filter(group==0) %>%
  select(feature) %>% unlist() %>% as.character()

length(cluster_markers_sig)
length(cluster_markers_presto_sig)
length(intersect(cluster_markers_sig, cluster_markers_presto_sig))
setdiff(cluster_markers_sig, cluster_markers_presto_sig)
setdiff(cluster_markers_presto_sig, cluster_markers_sig)

# Selecting top 8 markers for each cluster
cluster_markers_sig <- cluster_markers %>%
  filter(p_val_adj<0.01, abs(avg_log2FC)>0.5) %>%
  group_by(cluster) %>%
  slice_max(order_by = abs(avg_log2FC), n = 8) %>%
  ungroup %>% select(gene) %>% unlist() %>% as.character() %>% unique()

```

```{r, fig.height=8}
# Heatmap of top markers
# seurat_object = Seurat object with all features normalized and scaled
# plot_features = a vector a features to plot
# group_var = e.g. cluster
# group_colors = named vector of colors
# column_title = plot title
hm <- create_dotplot_heatmap(seurat_object = endo_mesen_reclustered,
                             plot_features = cluster_markers_sig,
                             group_var = "snn_res.0.8",
                             group_colors = cluster_col,
                             column_title = "Endothelial/Mesenchymal")

```

### Read cluster annotations from a Google Sheet

```{r}

endo_mesen_annot <- read_sheet(ct_annot, sheet = "Endo_Mesen")

```

### Adding annotations to the Seurat object and saving

```{r}

endo_mesen_reclustered$lineage <- mapvalues(x = endo_mesen_reclustered$snn_res.0.8,
                                            from = endo_mesen_annot$snn_res.0.8,
                                            to = endo_mesen_annot$lineage)

#saveRDS(endo_mesen_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_reclustered_annotated.rds")
endo_mesen_reclustered <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/endo_mesen_reclustered_annotated.rds")

```

### Visualizing

```{r}

DimPlot(endo_mesen_reclustered,
        group.by = "lineage",
        reduction = "umap",
        raster = T,
        cols = lineage_col,
        label = T,
        label.box = T,
        label.size = 3,
        repel = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()

```

### Subsetting endothelial and mesenchymal cells and reclustering

```{r}

mesenchymal <- subset(endo_mesen_reclustered, subset = lineage == "Mesenchymal")
mesenchymal_reclustered <- recluster(mesenchymal)

endothelial <- subset(endo_mesen_reclustered, subset = lineage == "Endothelial")
endothelial_reclustered <- recluster(endothelial)

```

### Annotating mesenchymal cells

```{r}
DimPlot(mesenchymal_reclustered,
        group.by = "snn_res.0.3",
        reduction = "umap",
        raster = T,
        cols = cluster_col,
        label = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()
```

```{r, fig.height=10}
# All mesenchymal markers
FeaturePlot(mesenchymal_reclustered,
            features = mesenchymal_features,
            ncol = 3,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()
```

```{r, fig.height=8}
DotPlot(mesenchymal_reclustered,
        features = mesenchymal_features,
        group.by = "snn_res.0.3",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()
```

### Cluster markers

```{r}
Idents(mesenchymal_reclustered) <- mesenchymal_reclustered$snn_res.0.3
cluster_markers <- FindAllMarkers(mesenchymal_reclustered,
                                  assay = "base_RNA",
                                  logfc.threshold = 0.25,
                                  test.use = "wilcox",
                                  slot = "data",
                                  min.pct = 0.1,
                                  verbose = F)

# Selecting top 8 markers for each cluster
cluster_markers_sig <- cluster_markers %>%
  filter(p_val_adj<0.01, abs(avg_log2FC)>0.5) %>%
  group_by(cluster) %>%
  slice_max(order_by = abs(avg_log2FC), n = 8) %>%
  ungroup %>% select(gene) %>% unlist() %>% as.character() %>% unique()

```

### Heatmap of top markers for mesenchymal clusters

```{r, fig.height=8}
hm <- create_dotplot_heatmap(seurat_object = mesenchymal_reclustered,
                             plot_features = cluster_markers_sig,
                             group_var = "snn_res.0.3",
                             group_colors = cluster_col,
                             column_title = "Mesenchymal")

```

### Adding annotations

```{r}

mesen_annot <- read_sheet(ct_annot, sheet = "Mesenchymal")

mesenchymal_reclustered$annotation_1 <- mapvalues(x = mesenchymal_reclustered$snn_res.0.3,
                                                  from = mesen_annot$snn_res.0.3,
                                                  to = mesen_annot$annotation_1)

DimPlot(mesenchymal_reclustered,
        group.by = "annotation_1",
        reduction = "umap",
        raster = T,
        cols = mesen_celltype_col,
        label = T,
        label.box = T,
        label.size = 3,
        repel = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()

```

### Annotating endothelial cells

```{r}

DimPlot(endothelial_reclustered,
        group.by = "snn_res.0.3",
        reduction = "umap",
        raster = T,
        cols = cluster_col,
        label = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()

```

```{r, fig.height=10}
# All endothelial markers
FeaturePlot(endothelial_reclustered,
            features = endothelial_features,
            ncol = 4,
            reduction = "umap",
            raster = T,
            cols = c("gray89", "tomato3")) &
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()
```

```{r, fig.height=8}

DotPlot(endothelial_reclustered,
        features = endothelial_features,
        group.by = "snn_res.0.3",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()

```

### Cluster markers

```{r}
Idents(endothelial_reclustered) <- endothelial_reclustered$snn_res.0.3
cluster_markers <- FindAllMarkers(endothelial_reclustered,
                                  assay = "base_RNA",
                                  logfc.threshold = 0.25,
                                  test.use = "wilcox",
                                  slot = "data",
                                  min.pct = 0.1,
                                  verbose = F)

# Selecting top 8 markers for each cluster
cluster_markers_sig <- cluster_markers %>%
  filter(p_val_adj<0.01, abs(avg_log2FC)>0.5) %>%
  group_by(cluster) %>%
  slice_max(order_by = abs(avg_log2FC), n = 8) %>%
  ungroup %>% select(gene) %>% unlist() %>% as.character() %>% unique()

```

### Heatmap of top markers for endothelial clusters

```{r, fig.height=8}
hm <- create_dotplot_heatmap(seurat_object = endothelial_reclustered,
                             plot_features = cluster_markers_sig,
                             group_var = "snn_res.0.3",
                             group_colors = cluster_col,
                             column_title = "Mesenchymal")

```

### Adding annotations

```{r}

endo_annot <- read_sheet(ct_annot, sheet = "Endothelial")

endothelial_reclustered$annotation_1 <- mapvalues(x = endothelial_reclustered$snn_res.0.3,
                                                  from = endo_annot$snn_res.0.3,
                                                  to = endo_annot$annotation_1)

DimPlot(endothelial_reclustered,
        group.by = "annotation_1",
        reduction = "umap",
        raster = T,
        cols = endo_celltype_col,
        label = T,
        label.box = T,
        label.size = 3,
        repel = T) +
  coord_fixed(ratio = 1) &
  theme_minimal() &
  NoLegend()

```

### Saving object

```{r}

#saveRDS(mesenchymal_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/mesenchymal_reclustered_annotated.rds")
#saveRDS(endothelial_reclustered, "/scratch/hnatri/ILD/ILD_spatial_ASE/endothelial_reclustered_annotated.rds")

```