---
title: "Post-clustering visualization"
author: "heinin"
date: "2023-12-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

### Packages and environment variables

```{r}

suppressPackageStartupMessages({library(cli)
                                library(Seurat)
                                library(SeuratObject)
                                library(SeuratDisk)
                                library(tidyverse)
                                library(tibble)
                                library(ggplot2)
                                library(ggpubr)
                                library(ggrepel)
                                library(workflowr)})

setwd("/home/hnatri/ILD_ASE_Xenium/")
set.seed(9999)
options(ggrepel.max.overlaps = Inf)

# Colors, themes, cell type markers, and plot functions
source("/home/hnatri/ILD_ASE_Xenium/code/colors_themes.R")
source("/home/hnatri/ILD_ASE_Xenium/code/plot_functions.R")

```


### Import data

```{r}

# Original clustering, n_pcs = 20
seurat_20 <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/ILDASE_seurat_2.rds")
DefaultAssay(seurat_20)

# n_pcs = 30
seurat_30 <- readRDS("/scratch/hnatri/ILD/ILD_spatial_ASE/ILDASE_seurat_PC-30_NCOMP-80.rds")

```

### Selecting the best clustering for downstream analysis

```{r}

p1 <- DimPlot(seurat_20,
        group.by = "leiden_res0.5",
        cols = cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 0.5) +
  theme_minimal() +
  NoLegend() +
  ggtitle("20 PCs")

p2 <- DimPlot(seurat_30,
        group.by = "leiden_res0.5",
        cols = cluster_col,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 0.5) +
  theme_minimal() +
  NoLegend() +
  ggtitle("30 PCs")
  

(p1 + p2)

# Using the reduction with 20 PCs
seurat_object <- seurat_20

rm(seurat_30)

```

### Adding spatial coordinates as a dimensionality reduction to the Seurat object

```{r}
position_xy <- cbind(seurat_object$x_centroid,
                     seurat_object$y_centroid)
row.names(position_xy) <- row.names(seurat_object@meta.data)
colnames(position_xy) <- c("SP_1", "SP_2")
seurat_object[["sp"]] <- CreateDimReducObject(embeddings = position_xy,
                                              key = "SP_",
                                              assay = DefaultAssay(seurat_object))

DimPlot(seurat_object,
        reduction = "sp",
        group.by = "sample",
        split.by = "sample",
        ncol = 5,
        #label = TRUE,
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend()

```

### Splitting the merged samples

MERGED_TILD010_VUHD115_TILD126

```{r}
unique(seurat_object$sample)[grep("MERGED", unique(seurat_object$sample))]

sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD010_VUHD115_TILD126")

MERGED_TILD010_VUHD115_TILD126 <- DimPlot(sample_subset,
                                          reduction = "sp",
                                          group.by = "sample",
                                          raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD010_VUHD115_TILD126")

# Finding coordinates to split the samples
MERGED_TILD010_VUHD115_TILD126 +
  # First sample
  geom_segment(aes(x = 400, y = 150, xend = 4300, yend = 150), linetype = "dashed") +
  geom_segment(aes(x = 400, y = 150, xend = 400, yend = 4300), linetype = "dashed") +
  geom_segment(aes(x = 400, y = 4300, xend = 4300, yend = 4300), linetype = "dashed") +
  geom_segment(aes(x = 4300, y = 4300, xend = 4300, yend = 150), linetype = "dashed") +
  # Second sample
  geom_segment(aes(x = 4900, y = 4100, xend = 4900, yend = 7900), linetype = "dashed", color = "forestgreen") +
  geom_segment(aes(x = 4900, y = 7900, xend = 8000, yend = 7900), linetype = "dashed", color = "forestgreen") +
  geom_segment(aes(x = 8000, y = 7900, xend = 8000, yend = 4100), linetype = "dashed", color = "forestgreen") +
  geom_segment(aes(x = 8000, y = 4100, xend = 4900, yend = 4100), linetype = "dashed", color = "forestgreen") +
  # Third sample
  geom_segment(aes(x = 4500, y = 180, xend = 4500, yend = 4000), linetype = "dashed", color = "tomato") +
  geom_segment(aes(x = 4500, y = 4000, xend = 8000, yend = 4000), linetype = "dashed", color = "tomato") +
  geom_segment(aes(x = 8000, y = 4000, xend = 8000, yend = 180), linetype = "dashed", color = "tomato") +
  geom_segment(aes(x = 8000, y = 180, xend = 4500, yend = 180), linetype = "dashed", color = "tomato")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: TILD126
lim.xleft <- 400
lim.xright <- 4300
lim.ybottom <- 150
lim.ytop <- 4300

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TIL010
lim.xleft <- 4900
lim.xright <- 8000
lim.ybottom <- 4100
lim.ytop <- 7900

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Sample 3: VUHD115
lim.xleft <- 4500
lim.xright <- 8000
lim.ybottom <- 180
lim.ytop <- 4000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample3 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample3 <- sample3

sample3_cells <- coord.pairs %>% filter(sample3 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "TILD126",
                                        ifelse(colnames(sample_subset) %in% sample2_cells, "TILD010",
                                               ifelse(colnames(sample_subset) %in% sample3_cells, "VUHD115",
                                                      "drop")))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("MERGED_TILD010_VUHD115_TILD126")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD010_VUHD115_TILD126 <- ifelse(colnames(seurat_object) %in% sample1_cells, "TILD126",
                                                ifelse(colnames(seurat_object) %in% sample2_cells, "TILD010",
                                                ifelse(colnames(seurat_object) %in% sample3_cells, "VUHD115",
                                                ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                                NA))))

```

MERGED_TILD028_TILD074

```{r}
sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD028_TILD074")

MERGED_TILD028_TILD074 <- DimPlot(sample_subset,
                                   reduction = "sp",
                                   group.by = "sample",
                                   raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD028_TILD074")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: TILD074
lim.xleft <- 0
lim.xright <- 3500
lim.ybottom <- 0
lim.ytop <- 3000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TILD028
lim.xleft <- 0
lim.xright <- 4000
lim.ybottom <- 3300
lim.ytop <- 8000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "TILD074",
                                        ifelse(colnames(sample_subset) %in% sample2_cells, "TILD028",
                                                      "drop"))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("MERGED_TILD028_TILD074")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD028_TILD074 <- ifelse(colnames(seurat_object) %in% sample1_cells, "TILD074",
                                        ifelse(colnames(seurat_object) %in% sample2_cells, "TILD028",
                                        ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                        NA)))

```

MERGED_TILD041_TILD055

```{r}
sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD041_TILD055")

MERGED_TILD041_TILD055 <- DimPlot(sample_subset,
                                   reduction = "sp",
                                   group.by = "sample",
                                   raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD041_TILD055")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: TILD005
lim.xleft <- 0
lim.xright <- 4200
lim.ybottom <- 0
lim.ytop <- 4500

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TILD041
lim.xleft <- 4500
lim.xright <- 8000
lim.ybottom <- 0
lim.ytop <- 4500

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "TILD055",
                                        ifelse(colnames(sample_subset) %in% sample2_cells, "TILD041",
                                                      "drop"))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("MERGED_TILD041_TILD055")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD041_TILD055 <- ifelse(colnames(seurat_object) %in% sample1_cells, "TILD055",
                                        ifelse(colnames(seurat_object) %in% sample2_cells, "TILD041",
                                        ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                        NA)))

```

MERGED_TILD049_TILD111_TILD080_VUHD113

```{r}
sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD049_TILD111_TILD080_VUHD113")

MERGED_TILD049_TILD111_TILD080_VUHD113 <- DimPlot(sample_subset,
                                                  reduction = "sp",
                                                  group.by = "sample",
                                                  raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD049_TILD111_TILD080_VUHD113")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: VUHD113
lim.xleft <- 0
lim.xright <- 3000
lim.ybottom <- 0
lim.ytop <- 600

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TILD080
lim.xleft <- 1000
lim.xright <- 5000
lim.ybottom <- 1200
lim.ytop <- 2800

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Sample 3: TILD111
lim.xleft <- 0
lim.xright <- 5000
lim.ybottom <- 3500
lim.ytop <- 5300

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample3 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample3 <- sample3

sample3_cells <- coord.pairs %>% filter(sample3 == "in") %>% rownames()

# Sample 4: TILD049
lim.xleft <- 0
lim.xright <- 5000
lim.ybottom <- 5800
lim.ytop <- 10100

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample4 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample4 <- sample4

sample4_cells <- coord.pairs %>% filter(sample4 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "VUHD113",
                                 ifelse(colnames(sample_subset) %in% sample2_cells, "TILD080",
                                 ifelse(colnames(sample_subset) %in% sample3_cells, "TILD111",
                                 ifelse(colnames(sample_subset) %in% sample4_cells, "TILD049",
                                 "drop"))))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("MERGED_TILD049_TILD111_TILD080_VUHD113")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD049_TILD111_TILD080_VUHD113<- ifelse(colnames(seurat_object) %in% sample1_cells, "VUHD113",
                                                              ifelse(colnames(seurat_object) %in% sample2_cells, "TILD080",
                                                              ifelse(colnames(seurat_object) %in% sample3_cells, "TILD111",
                                                              ifelse(colnames(seurat_object) %in% sample4_cells, "TILD049",
                                                              ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                                              NA)))))

```

MERGED_TILD059_TILD113

```{r}

sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD059_TILD113")

MERGED_TILD059_TILD113 <- DimPlot(sample_subset,
                                   reduction = "sp",
                                   group.by = "sample",
                                   raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD059_TILD113")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: TILD059
lim.xleft <- 0
lim.xright <- 5000
lim.ybottom <- 3200
lim.ytop <- 5000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TILD113
lim.xleft <- 0
lim.xright <- 3300
lim.ybottom <- 0
lim.ytop <- 2300

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "TILD059",
                                        ifelse(colnames(sample_subset) %in% sample2_cells, "TILD113",
                                               "drop"))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("MERGED_TILD059_TILD113")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD059_TILD113 <- ifelse(colnames(seurat_object) %in% sample1_cells, "TILD059",
                                        ifelse(colnames(seurat_object) %in% sample2_cells, "TILD113",
                                        ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                        NA)))

```

MERGED_TILD109_THD0016

```{r}

sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD109_THD0016")

MERGED_TILD109_THD0016 <- DimPlot(sample_subset,
                                   reduction = "sp",
                                   group.by = "sample",
                                   raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD109_THD0016")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: THD0016
lim.xleft <- 0
lim.xright <- 5000
lim.ybottom <- 0
lim.ytop <- 2600

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TILD109
lim.xleft <- 0
lim.xright <- 5000
lim.ybottom <- 3300
lim.ytop <- 6500

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "THD0016",
                                        ifelse(colnames(sample_subset) %in% sample2_cells, "TILD109",
                                               "drop"))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("x")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD109_THD0016 <- ifelse(colnames(seurat_object) %in% sample1_cells, "THD0016",
                                        ifelse(colnames(seurat_object) %in% sample2_cells, "TILD109",
                                        ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                        NA)))
```

MERGED_TILD136_VUILD102_TILD102_VUILD78

```{r}

sample_subset <- subset(seurat_object, subset = sample == "MERGED_TILD136_VUILD102_TILD102_VUILD78")

MERGED_TILD049_TILD111_TILD080_VUHD113 <- DimPlot(sample_subset,
                                                  reduction = "sp",
                                                  group.by = "sample",
                                                  raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  NoLegend() +
  ggtitle("MERGED_TILD136_VUILD102_TILD102_VUILD78")

# Adding sample names to match coordinates
coord.pairs <- sample_subset@meta.data[,c("x_centroid", "y_centroid")]
tolerance <- 0

# Sample 1: VUILD78
lim.xleft <- 0
lim.xright <- 3200
lim.ybottom <- 0
lim.ytop <- 3000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample1 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample1 <- sample1

sample1_cells <- coord.pairs %>% filter(sample1 == "in") %>% rownames()

# Sample 2: TILD102
lim.xleft <- 0
lim.xright <- 3200
lim.ybottom <- 3200
lim.ytop <- 4000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample2 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample2 <- sample2

sample2_cells <- coord.pairs %>% filter(sample2 == "in") %>% rownames()

# Sample 3: VUILD102
lim.xleft <- 0
lim.xright <- 3200
lim.ybottom <- 5500
lim.ytop <- 8000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample3 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample3 <- sample3

sample3_cells <- coord.pairs %>% filter(sample3 == "in") %>% rownames()

# Sample 4: TILD136
lim.xleft <- 0
lim.xright <- 3200
lim.ybottom <- 8500
lim.ytop <- 11000

exclude.xleft <- lim.xleft + tolerance
exclude.xright <- lim.xright - tolerance
exclude.ybottom <- lim.ybottom + tolerance
exclude.ytop <- lim.ytop - tolerance
sample4 <- c("out", "in")[1+(findInterval(coord.pairs[ , 1], c(exclude.xleft, exclude.xright))==1 &
                           findInterval(coord.pairs[ , 2], c(exclude.ybottom, exclude.ytop))==1)]
coord.pairs$sample4 <- sample4

sample4_cells <- coord.pairs %>% filter(sample4 == "in") %>% rownames()

# Adding sample names to the Seurat object subset
sample_subset$unmerged_sample <- ifelse(colnames(sample_subset) %in% sample1_cells, "VUILD78",
                                 ifelse(colnames(sample_subset) %in% sample2_cells, "TILD102",
                                 ifelse(colnames(sample_subset) %in% sample3_cells, "VUILD102",
                                 ifelse(colnames(sample_subset) %in% sample4_cells, "TILD136",
                                 "drop"))))

DimPlot(sample_subset,
        reduction = "sp",
        group.by = "unmerged_sample",
        raster = T) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  #NoLegend() +
  ggtitle("MERGED_TILD136_VUILD102_TILD102_VUILD78")

# Adding sample names to the complete Seurat object
seurat_object$MERGED_TILD136_VUILD102_TILD102_VUILD78 <- ifelse(colnames(seurat_object) %in% sample1_cells, "VUILD78",
                                                                ifelse(colnames(seurat_object) %in% sample2_cells, "TILD102",
                                                                ifelse(colnames(seurat_object) %in% sample3_cells, "VUILD102",
                                                                ifelse(colnames(seurat_object) %in% sample4_cells, "TILD136",
                                                                ifelse(colnames(seurat_object) %in% colnames(sample_subset), "drop",
                                                                NA)))))
```

#### Combining sample name columns

```{r}
metadata <- seurat_object@meta.data %>%
  mutate(merged_sample = coalesce(MERGED_TILD010_VUHD115_TILD126,
                                  MERGED_TILD028_TILD074,
                                  MERGED_TILD041_TILD055,
                                  MERGED_TILD049_TILD111_TILD080_VUHD113,
                                  MERGED_TILD059_TILD113,
                                  MERGED_TILD109_THD0016,
                                  MERGED_TILD136_VUILD102_TILD102_VUILD78))

table(metadata$sample)
table(metadata$merged_sample)

metadata <- metadata %>% mutate(Sample_Name = coalesce(merged_sample, sample))
table(metadata$Sample_Name)
seurat_object@meta.data <- metadata

# Dropping cells at the margins of the merged samples
seurat_object <- subset(seurat_object,
                        subset = Sample_Name == "drop",
                        invert = T)

# Adding sample type information
seurat_object$Sample_Type <- ifelse(grepl("ILD", seurat_object$Sample_Name), "ILD", "HD")
table(seurat_object$Sample_Type)

# Saving the object to lab storage
#saveRDS(seurat_object, "/tgen_labs/banovich/IPF/Spatial_ASE/ILD_ASE_Xenium_processed_npcs20.rds")
```

### Cell numbers per sample

```{r}

table(seurat_object$Sample_Name) %>% as.data.frame() %>%
  rename(Sample_Name = Var1,
         n_cells = Freq) %>%
  ggplot(aes(x = reorder(Sample_Name, -n_cells), y = n_cells, fill = Sample_Name)) + 
  geom_bar(position="stack", stat="identity") + 
  scale_fill_manual(values = sample_col) +
  theme_bw() + 
  xlab("") + 
  ylab("# cells") + 
  NoLegend() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=0.9))

```

### DimPlots

```{r}

DimPlot(seurat_object,
        group.by = "leiden_res0.5",
        cols = cluster_col,
        #split.by = "sample",
        #ncol = 5,
        reduction = "umap",
        raster = T,
        label = T) +
  coord_fixed(ratio = 1) +
  NoLegend() +
  theme_minimal()

DimPlot(seurat_object,
        group.by = "leiden_res0.5",
        cols = cluster_col,
        split.by = "sample",
        ncol = 8,
        reduction = "umap",
        raster = T,
        label = F) +
  coord_fixed(ratio = 1) +
  theme_minimal()

DimPlot(seurat_object,
        group.by = "leiden_res0.5",
        cols = cluster_col,
        split.by = "Sample_Name",
        ncol = 8,
        reduction = "umap",
        raster = T,
        label = F) +
  coord_fixed(ratio = 1) +
  theme_minimal()

```

### Feature expression

```{r, fig.height=12}

DotPlot(seurat_object,
        features = unique(c(epithelial_features,
                            immune_features,
                            endothelial_features,
                            mesenchymal_features)),
        group.by = "leiden_res0.5",
        cols = c("azure", "tomato3")) +
  coord_flip() +
  theme_minimal()

```

### Cell type proportions by sample and sample type

```{r}
# Scatter plot of cell type proportions by sample type

# seurat_object = Seurat object
# group_var = e.g. CD3_status
# group1 = e.g. High
# group2 = e.g. Low
# plot_var = e.g. celltype
# plot_colors = A named vector of colors to use in the plot, corresponding to plot_var (e.g. cell type colors)
# var_names = Used as axis titles
# legend_title = Legend title, ("" for no title)
create_clusterpropplot(seurat_object = seurat_object,
                       group_var = "Sample_Type",
                       group1 = "HD",
                       group2 = "ILD",
                       plot_var = "leiden_res0.5",
                       plot_colors = cluster_col,
                       var_names = c("HD", "ILD"),
                       legend_title = "")

# Barplot

# seurat_object = Seurat object
# plot_var = e.g. cluster
# group_var = e.g. response (cell proportions are plotted for each group)
# group_levels = a vector for ordering the grouping variable levels
# plot_levels = a vector for ordering the plot variable levels
# plot_colors = A named vector of colors to use in the plot, corresponding to plot_var (e.g. cluster colors)
# var_names = Used as axis titles, c("plot_var_name", "group_var_name")
# legend_title = Legend title, ("" for no title)
create_barplot(seurat_object = seurat_object,
               group_var = "Sample_Type",
               plot_var = "leiden_res0.5",
               group_levels = c("ILD", "HD"),
               plot_levels = names(cluster_col),
               plot_colors = cluster_col,
               var_names = c("Cluster", "Sample Type"),
               legend_title = "")

# Stacked barplot of cell type proportions across samples
create_barplot(seurat_object = seurat_object,
               group_var = "Sample_Name",
               plot_var = "leiden_res0.5",
               group_levels = sort(names(sample_col)),
               plot_levels = names(cluster_col),
               plot_colors = cluster_col,
               var_names = c("Cluster", "Sample"),
               legend_title = "") +
  coord_flip()

```



